Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package

### 2020-09-18 020-simulation-algorithm.Rmd
update case study with new syntax
get started after a long break and note the current limitations

- Discrete margins pose some difficulties and matching *unadjusted* correlation
  measures is not exact (empirically approximate).
- Exact Pearson solutions are possible but require computing $d \choose 2$
   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
- We GPU accelerate whenever possible during our proposed High-Performance NORTA
  algorithm.
- One thorny problem when dealing with high-dimensional correlation matrices is
they can become non-positive definite through either during estimation or
bivariate transformations.
- When converting nonparametric correlations to Pearson (first step above) the
  resultant correlation may not be positive definite (PD).
- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
- In practice, this loss in accuracy typically has little impact on performance,
  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).

test workflow again

>>>>

- discuss the consequences of the transformations to obtain the corresponding Pearson matrix and describe the quadratic algorithm implemented to address this (Alex).
- Discrete margins pose some difficulties and matching *unadjusted* correlation
  measures is not exact (empirically approximate).
- Exact Pearson solutions are possible but require computing $d \choose 2$
   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
- We GPU accelerate certain steps (Alex which steps?) possible during our proposed High-Performance NORTA
  algorithm.
- One thorny problem when dealing with high-dimensional correlation matrices is
they can become non-positive definite through either during estimation or
bivariate transformations.
- When converting nonparametric correlations to Pearson (first step above) the
  resultant correlation may not be positive definite (PD).
- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
- In practice, this loss in accuracy typically has little impact on performance,
  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).

test workflow again
test after a long while

### 2020-09-18 030-bigsimr-package.Rmd
get started after a long break and note the current limitations

```{r setup, eval = TRUE}
## setwd('~/Research/article_bigsimr/'); bookdown::preview_chapter('030-bigsimr-package.Rmd')
## bookdown::render_book('index.html')
library(bigsimr)
## reticulate::use_condaenv( 'bigsimr-cpu' )
```

@@

```{r}
margins = alist(
  qnorm(mean = 3.14, sd = 0.1),
  qbeta(shape1 = 1, shape2 = 4),
  qnbinom(size = 10, prob = 0.75)
)
```

@@

```{r}
rho <- matrix(0.5, nrow = 3, ncol = 3)
diag(rho) <- 1.0
rho
```

@@

```{r}
x <- rvec(10, rho = rho, margins = margins, type = "spearman")
```

```{r, echo=FALSE}
warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
```

@@

```{r}
x
```

@@

```{r, fig.width=7}
x <- rvec(1e6, rho = rho, margins = margins, type = "spearman")

@@

```{r}
bigsimr::fastCor(x, method = 'spearman')
```

@@

```{r, computeBounds, echo = TRUE, eval=FALSE}
## Possibly bug? or configuration
## Error: segfault from C stack overflow on my Okapi bigsimr env 
rho_bounds <- bigsimr::computeCorBounds(margins = margins, type = 'spearman', cores = 1, reps = 1e5)
rho_bounds
```

@@

```{r exampleRslurm, echo=FALSE, eval=FALSE, include=FALSE}
## run this code once and use the generated rds later.
## cyberhelp.sesync.org/rslurm/articles/rslurm.html
test_func <- function(par_mu, par_sd) {
    samp <- rnorm(10^6, par_mu, par_sd)
    c(s_mu = mean(samp), s_sd = sd(samp))
}
pars <- data.frame(par_mu = 1:10,
                   par_sd = seq(0.1, 1, length.out = 10))
library(rslurm)
sjob <- slurm_apply(test_func, pars, jobname = 'test_apply',
                    nodes = 2, cpus_per_node = 2, submit = TRUE)
list.files('_rslurm_test_apply', 'results')
res <- get_slurm_out(sjob, outtype = 'table')
head(res, 3)
cleanup_files(sjob) ## careful this deletes output also

## single function evaluation
sjob <- slurm_call(test_func, jobname = 'test_call',
                   list(par_mu = 5, par_sd = 1), submit = FALSE)

## try rvec
sjob <- slurm_call(rvec, jobname = 'rvec',
                   list(par_mu = 5, par_sd = 1), submit = FALSE)
```

@@

```{r rvecRslurmRes, echo=TRUE, eval=TRUE}
res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
head(res, 3)
## cleanup_files(sjob) ## careful this deletes output also
```

>>>>

```{r setup3, eval = TRUE}
## setwd('~/Research/article_bigsimr/'); bookdown::preview_chapter('030-bigsimr-package.Rmd')
## bookdown::render_book('index.html')
library(bigsimr)
library(reticulate)
reticulate::use_condaenv("bigsimr-cpu")
```

@@

```{r setMargins}
margins = alist(
  qnorm(mean = 3.14, sd = 0.1),
  qbeta(shape1 = 1, shape2 = 4),
  qnbinom(size = 10, prob = 0.75)
)
```

@@

```{r setCor}
rho <- matrix(0.5, nrow = 3, ncol = 3)
diag(rho) <- 1.0
rho
```

@@

```{r rvec10}
x <- bigsimr::rvec(10, rho = rho, margins = margins, type = "spearman")
```

```{r gpuWarn, echo=FALSE}
warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
```

@@

```{r printX}
x
```

@@

```{r rvec1e6, fig.width=7}
x <- rvec(1e6, rho = rho, margins = margins, type = "spearman")

@@

```{r fastCor}
bigsimr::fastCor(x, method = 'spearman')
```

@@

```{r computeBounds, echo = TRUE, eval=FALSE}
## Possibly bug? or configuration
## Error: segfault from C stack overflow on my Okapi bigsimr env 
rho_bounds <- bigsimr::computeCorBounds(margins = margins, type = 'spearman', cores = 1, reps = 1e5)
rho_bounds
```

@@

@@

```{r rvecRslurmRes, echo=TRUE, eval=FALSE}
res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
head(res, 3)
## cleanup_files(sjob) ## careful this deletes output also
```

### 2020-09-18 050-some-applications.Rmd
get started after a long break and note the current limitations

```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
library(bigsimr)
reticulate::use_condaenv("py37")
CORES <- parallel::detectCores() - 1
set.seed(06082020)
## devtools::install_github("ggobi/GGally")
## devtools::install_github( "adknudson/bigsimr")
## A conda environment is also activitied (see README)
library(bigsimr)
library(GGally)
```

>>>>

```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
library(bigsimr)
reticulate::use_condaenv("bigsimr")
CORES <- parallel::detectCores() - 1
set.seed(06082020)
## devtools::install_github("ggobi/GGally")
## devtools::install_github( "adknudson/bigsimr")
## A conda environment is also activitied (see README)
library(bigsimr)
library(GGally)
```

### 2020-09-18 060-conclusion-discussion.Rmd
migrate to html based preview
get started after a long break and note the current limitations

There are of course limitations to the methodology and implementation. The algorithm requires invertible marginal cdfs. This leaves some restriction to the available marginal probability distributions (DOES IT? FOR EXAMPLE?) An additional source of error is after converting dependencies (step i; ${\bf R_{Kendall}})$) the resultant matrix may not be positive definite (PD). Then the nearest PD matrix is used. From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package. A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment. The algorithm does amends itself well to parallelization and graphical processing unit acceleration [@Li2019gpu] and advanced users may take advantage of those techniques.

Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly. Further, more sophisticated approaches could involve simulation algorithms that are "estimation aware" and so could explore the role of covariance estimation within the simulation. Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models. On the implementation side, employing graphical process unit acceleration could produce substantial speedups over our multi-core approach.

>>>>

Mention `rslurm`

There are of course limitations to the methodology and implementation. The most obvious missing feature the proposed methodology is the inability to match a Pearson correlation matrix exactly. As discussed above, this is a computational intense procedure and not a natural choice for the posed problem (seeking to simulate non-normal margins in a scaleable fashion). Yet it is an important aspect that paralleziable, approximation based apporaches are promising [ref Hermite polynomials]. Further, while we provide the ability of the user to specify discrete marginals, exact matching a desired dependency measure is not yet obtained. One potential solution for this is *rescale* to account for ties [ref]. Finally we note, that the algorithm requires invertible marginal cdfs. This gives some restriction to the available marginal probability distributions (DOES IT? FOR EXAMPLE?).

From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package. A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment. The algorithm does amends itself well to parallelization and graphical processing unit acceleration [@Li2019gpu] and advanced users may take advantage of those techniques.

Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly and more methods developed specifically for discrete distributions. Further, more sophisticated approaches could involve simulation algorithms that are "estimation aware" and so could explore the role of covariance estimation within the simulation. Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models. On the implementation side, employing graphical process unit acceleration could produce substantial speedups over our multi-core approach.

### 2020-09-18 index.Rmd
get started after a long break and note the current limitations

--- 
title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
author: [A.G. Schissler, A. Knudson, T. Kozubowski, A. Panorksa]
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
site: bookdown::bookdown_site
bibliography: [bigsimr.bib, packages.bib]
biblio-style: apalike
link-citations: yes
abstract: |
  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the `bigsimr` R package.
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
set.seed(06082020)
```

```{bash copyBib, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
```

@@

@@

Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest. Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018]. RNA-sequencing analysis has garnered much interest in the statistical literature (refs). Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes. In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].

@@

With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):

>>>>

--- 
title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
author: [A.G. Schissler, R.D.L. Foote, A.D. Knudson, T.J. Kozubowski, A.K. Panorska]
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
site: bookdown::bookdown_site
bibliography: [bigsimr.bib, packages.bib]
biblio-style: apalike
link-citations: yes
abstract: |
  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the `bigsimr` R package.
---

```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
set.seed(06082020)
```

```{r copyBib, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
system( 'cp ~/bib/bigsimr.bib ./.' )
```

@@

```{r serveHelper, include=FALSE, eval=FALSE}
## rmote::start_rmote()
## library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
bookdown::serve_book()
```

```{r countWords, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
rmdFiles <- dir(pattern = 'Rmd$', ignore.case= T)
## rmdFiles <- rmdFiles[-grep('^_', rmdFiles)]
rmdFiles <- rmdFiles[-grep('references', rmdFiles)]
## rmdFiles <- rmdFiles[-grep('README', rmdFiles)]
## rmdFiles <- rmdFiles[-grep('article', rmdFiles)] ## after generating a single document
## wordcountaddin::word_count( rmdFiles[11] )
(counts <- sapply( rmdFiles, wordcountaddin::word_count, simplify = T ))
sum(counts)
```

@@

Along those lines, simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest. Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018]. RNA-sequencing analysis has garnered much interest in the statistical literature (refs). Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes. In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].

@@

With our application in mind and seeking a fairly general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):
